---
title: "Homework 0"
author: "Bar Hana Yehezkel 322723206"
description: "The goal of this homework is to introduce you to R, RStudio and R Markdown. In the assignment you will need to submit the html file created by knitting this .rmd file, where every code chunk is properly filled based on the instructions."
output: html_document
---



### Instructions

1. All assignments are individual! 
2. No submission for this assignment.
3. Questions about the assignment should be asked using the Assignments forums. 


The goal of this homework is to introduce you to R, RStudio and R Markdown. In the assignment you will need to submit the .rmd file created by this markdown project, where every chunk is filled properly based on the instructions.

```{r, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
### 'Questions' 
Q1. Given a vector x = (17, 14, 4, 5, 13, 12, 10) set all elements of this vector that are greater than 10 to be equal to 4.  (10 points)

```{r, eval = FALSE}
x <- c(17, 14, 4, 5, 13, 12, 10)
x[x > 10] <- 4
x
```


Q2. Create a function that receives a vector and returns the max, min, mean values.  (10 points)

```{r, eval = FALSE}
myFunc <- function(vec) {

    return_list <- list(max = max(vec), min = min(vec), mean = mean(vec))
    return (return_list)
}
myFunc(c(67,2,4,5,23,1,54,6,7,8,33))

```

Q3. Create a function that calculates the factorial of an integer (n!) using a while loop.  (20 points)

```{r, eval = FALSE}
factorial <- function(n) {
  result <- 1
  while(n > 0) {
    result <- result * n
    n <- n - 1
  }
  return(result)
}
factorial(5) 
```

## Seatbelts dataset

Let's take a look at the seatbelts dataset, which is based on the "Road Casualties in Great Britain 1969-84" inbuilt R data set.
The 'Seatbelts' dataset in R is a multiple time-series that was commissioned by the Department of Transport in 1984 to measure differences in deaths before and after front seatbelt legislation was introduced on 31st January 1983.

For simpler view, run the following chunk:

```{r}
data(Seatbelts)
Seatbelts <- data.frame(
  Year=floor(time(Seatbelts)),
  Month=factor(cycle(Seatbelts), labels=month.abb), 
  Seatbelts)
```

```{r}
head(Seatbelts)
```

```{r}
str(Seatbelts)
```

Q4. sort and print 'drivers' variable. (10 points)
```{r, eval=FALSE}
sort(Seatbelts$drivers)
```

Q5. print rows 3-6 from Seatbelts dataset. (10 points)

```{r, eval=FALSE}
Seatbelts[3:6, ]
```

Q6. print the data for the year 1970 (10 points)
```{r, eval=FALSE}
Seatbelts[Seatbelts$Year == 1970, ]
```

Q7. calculate the mean of 'kms' attribute for January ('Jan') and for February ('Feb') months. (20 points)
```{r, eval=FALSE}
jan_set = Seatbelts[Seatbelts$Month == "Jan", ]
feb_set = Seatbelts[Seatbelts$Month == "Feb", ]

mean(jan_set$kms)
mean(feb_set$kms)
```

Q8. Use the 'boxplot' function (part of the R basic 'graphics' package, not ggplot) to create a boxplot for the 'kms' attribute for January ('Jan') data and another plot for the February ('Feb') months. (10 points)
```{r, eval=FALSE}
boxplot(jan_set$kms, main="Kms in January")
boxplot(feb_set$kms, main="Kms in February")
```

